---
title: "Lab 4a"
author: "Margaret Hilliard"
date: "9/21/2020"
output: html_document
bibliography: /Users/margarethilliard/Desktop/Evolutionary_Genomics/genomics-course/data/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, eval = FALSE)
```
load the library and read in the data 
```{r}
library("tidyverse")
surveys_complete <- read_csv("data/surveys_complete.csv")
```
# Plotting with ggplot2 
basic template is ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
use the ggplot() function and bind the plot to a specific data frame using the data argument
```{r}
ggplot(data = surveys_complete)
```
define an aesthetic mapping (using the aesthetic (aes) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.
```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
```
add ‘geoms’ – graphical representations of the data in the plot (points, lines, bars). ggplot2 offers many different geoms; we will use some common ones today, including:
        geom_point() for scatter plots, dot plots, etc.
        geom_boxplot() for, well, boxplots!
        geom_line() for trend lines, time series, etc.

To add a geom to the plot use + operator. Because we have two continuous variables, let’s use geom_point() first:
```{r}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```
Alternitively, assign plot to a variable
```{r}
surveys_plot <- ggplot(data = surveys_complete, 
                       mapping = aes(x = weight, y = hindfoot_length))
# Draw the plot
surveys_plot + 
    geom_point()
```
Then, we start modifying this plot to extract more information from it. For instance, we can add transparency (alpha) to avoid overplotting:
```{r}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1)
```
We can also add colors for all the points:
```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, color = "blue")
```
Or to color each species in the plot differently, you could use a vector as an input to the argument color. ggplot2 will provide a different color corresponding to different values in the vector. Here is an example where we color with species_id:
```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, aes(color = species_id))
```
# Challenge 1 
Use what you just learned to create a scatter plot of weight over species_id with the plot types showing in different colors. Is this a good way to show this type of data?
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_point(aes(color = plot_type))
```
# Boxplot
We can use boxplots to visualize the distribution of weight within each species:
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot()
```
By adding points to the boxplot, we can have a better idea of the number of measurements and of their distribution:
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.3, color = "tomato")
```
Notice how  the boxplot layer is behind the jitter layer? What do you need to change in the code to put the boxplot in front of the points such that it’s not hidden?
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_boxplot(alpha = 0) # change the order in which you're layering the components of the plot 
```
# Challenge 2 
Boxplots are useful summaries, but hide the shape of the distribution. For example, if there is a bimodal distribution, it would not be observed with a boxplot. An alternative to the boxplot is the violin plot (sometimes known as a beanplot), where the shape (of the density of points) is drawn.

Replace the box plot with a violin plot; see geom_violin().
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_violin()
```
In many types of data, it is important to consider the scale of the observations. For example, it may be worth changing the scale of the axis to better distribute the observations in the space of the plot. Changing the scale of the axes is done similarly to adding/modifying other components (i.e., by incrementally adding commands). Try making these modifications:

Represent weight on the log10 scale; see scale_y_log10()
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_violin() +
  scale_y_log10()
```
So far, we’ve looked at the distribution of weight within species. Try making a new plot to explore the distribution of another variable within each species.

Create boxplot for hindfoot_length. Overlay the boxplot layer on a jitter layer to show actual measurements.
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_boxplot()
  
# or 
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_violin()
```
Add color to the data points on your boxplot according to the plot from which the sample was taken (plot_id)
```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.3, aes(color = plot_id)) +
  geom_violin()
```
Hint: Check the class for plot_id. Consider changing the class of plot_id from integer to factor. Why does this change how R makes the graph?
```{r}
class(surveys_complete$plot_id)
as.factor(surveys_complete$plot_id)
```

```{r}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.3, aes(color = as.factor(surveys_complete$plot_id))) +
  geom_violin()
```
# Plotting time series data
Let’s calculate number of counts per year for each genus. First we need to group the data and count records within each group:
```{r}
yearly_counts <- surveys_complete %>%
  count(year, genus)
```
Timelapse data can be visualized as a line plot with years on the x-axis and counts on the y-axis:
```{r}
ggplot(data = yearly_counts, aes(x = year, y = n)) +
     geom_line()
```
Unfortunately, this does not work because we plotted data for all the genera together. We need to tell ggplot to draw a line for each genus by modifying the aesthetic function to include group = genus:
```{r}
ggplot(data = yearly_counts, aes(x = year, y = n, group = genus)) +
    geom_line()
```
We will be able to distinguish species in the plot if we add colors (using color also automatically groups the data):
```{r}
ggplot(data = yearly_counts, aes(x = year, y = n, color = genus)) +
    geom_line()
```
# Integrating the pipe operator with ggplot2
In the previous lesson, we saw how to use the pipe operator %>% to use different functions in a sequence and create a coherent workflow. We can also use the pipe operator to pass the data argument to the ggplot() function. The hard part is to remember that to build your ggplot, you need to use + and not %>%.
```{r}
yearly_counts %>% 
    ggplot(mapping = aes(x = year, y = n, color = genus)) +
    geom_line()
```
The pipe operator can also be used to link data manipulation with consequent data visualization.
```{r}
yearly_counts_graph <- surveys_complete %>%
    count(year, genus) %>% 
    ggplot(mapping = aes(x = year, y = n, color = genus)) +
    geom_line()

yearly_counts_graph
```
# Faceting
ggplot has a special technique called faceting that allows the user to split one plot into multiple plots based on a factor included in the dataset. We will use it to make a time series plot for each species:
```{r}
ggplot(data = yearly_counts, aes(x = year, y = n)) +
    geom_line() +
    facet_wrap(facets = vars(genus))
```
Now we would like to split the line in each plot by the sex of each individual measured. To do that we need to make counts in the data frame grouped by year, genus, and sex:
```{r}
yearly_sex_counts <- surveys_complete %>%
                      count(year, genus, sex)
```
We can now make the faceted plot by splitting further by sex using color (within a single plot):
```{r}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
  geom_line() +
  facet_wrap(facets =  vars(genus))
```
We can also facet both by sex and genus:
```{r}
ggplot(data = yearly_sex_counts, 
       mapping = aes(x = year, y = n, color = sex)) +
  geom_line() +
  facet_grid(rows = vars(sex), cols =  vars(genus))
```
You can also organise the panels only by rows (or only by columns):
```{r}
# One column, facet by rows
ggplot(data = yearly_sex_counts, 
       mapping = aes(x = year, y = n, color = sex)) +
  geom_line() +
  facet_grid(rows = vars(genus))
```

```{r}
# One row, facet by column
ggplot(data = yearly_sex_counts, 
       mapping = aes(x = year, y = n, color = sex)) +
  geom_line() +
  facet_grid(cols = vars(genus))
```
# ggplot2 themes
Usually plots with white background look more readable when printed. Every single component of a ggplot graph can be customized using the generic theme() function, as we will see below. However, there are pre-loaded themes available that change the overall appearance of the graph without much effort.

For example, we can change our previous graph to have a simpler white background using the theme_bw() function:
```{r}
 ggplot(data = yearly_sex_counts, 
        mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(vars(genus)) +
     theme_bw()
```
In addition to theme_bw(), which changes the plot background to white, ggplot2 comes with several other themes which can be useful to quickly change the look of your visualization. The complete list of themes is available at https://ggplot2.tidyverse.org/reference/ggtheme.html. theme_minimal() and theme_light() are popular, and theme_void() can be useful as a starting point to create a new hand-crafted theme.

The ggthemes package provides a wide variety of options.
# Challenge 3 
Use what you just learned to create a plot that depicts how the average weight of each species changes through the years.
```{r}
# summarize data by average weight of each species, separated by year 
yearly_weight <- surveys_complete %>%
                group_by(year, species_id) %>%
                 summarize(avg_weight = mean(weight))
```
```{r}
# plot by year and average weight, split by species 
ggplot(data = yearly_weight, mapping = aes(x=year, y=avg_weight)) +
   geom_line() +
   facet_wrap(vars(species_id)) +
   theme_bw()
```
# Customization
Take a look at the ggplot2 cheat sheet, and think of ways you could improve the plot.

Now, let’s change names of axes to something more informative than ‘year’ and ‘n’ and add a title to the figure:
```{r}
ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(vars(genus)) +
    labs(title = "Observed genera through time",
         x = "Year of observation",
         y = "Number of individuals") +
    theme_bw()
```
The axes have more informative names, but their readability can be improved by increasing the font size. This can be done with the generic theme() function:
```{r}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(vars(genus)) +
    labs(title = "Observed genera through time",
        x = "Year of observation",
        y = "Number of individuals") +
    theme_bw() +
    theme(text=element_text(size = 16))
```
After our manipulations, you may notice that the values on the x-axis are still not properly readable. Let’s change the orientation of the labels and adjust them vertically and horizontally so they don’t overlap. You can use a 90 degree angle, or experiment to find the appropriate angle for diagonally oriented labels. We can also modify the facet label text (strip.text) to italicize the genus names:
```{r}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(vars(genus)) +
    labs(title = "Observed genera through time",
        x = "Year of observation",
        y = "Number of individuals") +
    theme_bw() +
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 90, hjust = 0.5, vjust = 0.5),
                        axis.text.y = element_text(colour = "grey20", size = 12),
                        strip.text = element_text(face = "italic"),
                        text = element_text(size = 16))
```
If you like the changes you created better than the default theme, you can save them as an object to be able to easily apply them to other plots you may create:
```{r}
grey_theme <- theme(axis.text.x = element_text(colour="grey20", size = 12, 
                                               angle = 90, hjust = 0.5, 
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text=element_text(size = 16))

ggplot(surveys_complete, aes(x = species_id, y = hindfoot_length)) +
    geom_boxplot() +
    grey_theme
```
# Challenge 4 
With all of this information in hand, please take another five minutes to either improve one of the plots generated in this exercise or create a beautiful graph of your own. Use the RStudio ggplot2 cheat sheet for inspiration.
    Here are some ideas:
      * See if you can change the thickness of the lines.
      * Can you find a way to change the name of the legend? What about its labels?
      * Try using a different color palette (see http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).
```{r}
# From challenege 3 -- summarize data by average weight of each species, separated by year 
yearly_weight <- surveys_complete %>%
                group_by(year, species_id) %>%
                 summarize(avg_weight = mean(weight))
# plot by year and average weight, split by species 
ggplot(data = yearly_weight, mapping = aes(x=year, y=avg_weight)) +
   geom_line() +
   facet_wrap(vars(species_id)) +
   theme_bw()

#change axis labels 
ggplot(data = yearly_weight, mapping = aes(x=year, y=avg_weight)) +
   geom_line() +
   facet_wrap(vars(species_id)) +
  labs(title = "Species weight over time",
         x = "Year of observation",
         y = "Average weight of individuals") +
   theme_bw()

# easier to manipulate 
e <- ggplot(yearly_weight, aes(year, avg_weight))
e +
  geom_line () +
  facet_wrap(vars(species_id)) +
  labs(title = "Species weight over time",
         x = "Year of observation",
         y = "Average weight of individuals") 
  
```
Another example:
```{r}
w <-   ggplot(data = yearly_sex_counts, 
       mapping = aes(x = year, y = n, color = sex))
w +
  geom_line() +
  facet_grid(rows = vars(sex), cols =  vars(genus)) +
  theme(axis.text.x = element_text(color = "grey20", size = 8, angle = 90)) + # adjusting appearance of years
  labs(title = "Species counts by sex over time", # title and axis labels 
         x = "Year of observation",
         y = "Individuals observed") +
   theme(text=element_text(size = 9)) # improve readability 

```

Application written in R [@RCoreTeam] using the Shiny framework [@Chang2015]. 

# References 


