---
title: "Lab 5"
author: "Margaret Hilliard"
date: "9/21/2020"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, eval = FALSE)
```
# Joining tables
load the tidyverse library 
```{r}
library(tidyverse)
```
read in march data (filter to be just the US)
```{r}
Confirmed_State_3_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-13-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State") %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed))
```
check the structure of data 
```{r}
str(Confirmed_State_3_13)
```
read in september data (filter to be just the US)
```{r}
Confirmed_State_9_13 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed))
```
check the structure
```{r}
str(Confirmed_State_9_13)
```
Notice from the above `str’ calls that there are 53 states in the table on 3/13/2020 and 58 states in the table on 9/13/2020
```{r}
setdiff(Confirmed_State_9_13$Province_State, Confirmed_State_3_13$Province_State)
```
There are 4 US territories a field you may have notice last week “Recovered”. We don’t need the “Recovered” row so let’s filter it from the data set
```{r}
Confirmed_State_9_13 <- Confirmed_State_9_13 %>% 
  filter(Province_State != "Recovered") #taking the "recovered" column out of the data
```
Since all of the states and territories in 9/13/2020 are found in 3/13/2020 a left_join or full_join can be used to merge the tables
```{r}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_3_13_9_13_joined) #joining the two tables by province/state
```
This results in the values for Guam, Northern Mariana Islands, Puerto Rico and the Virgin Islands being NA for 3/13
```{r}
tail(Confirmed_State_3_13_9_13_joined, 5)
# or 
which(is.na(Confirmed_State_3_13_9_13_joined))
```
Renaming columns: Let’s turn the NAs into 0. Also notice that we know have columns Confirmed.x Country_Region.x, Confirmed.y Country_Region.y. So let’s rename the Confirmed columns and delete the Country_Region columns
```{r}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State")) %>% 
      rename(Confirmed_3_13_2020 = "Confirmed.x", Confirmed_9_13_2020 = "Confirmed.y") %>% 
      select(-Country_Region.x, -Country_Region.y) %>% 
      replace_na(list(Confirmed_3_13_2020 = 0))

head(Confirmed_State_3_13_9_13_joined)
```
check if NAs are now 0 
```{r}
which(is.na(Confirmed_State_3_13_9_13_joined))
```
# Switching between wide and long table formats
gather and spead were used in the Data Carpentries tutorial to switch between wide and long table formats. They work fine, but are being replaced by the newer functions pivot_longer and pivot_wider. which we will use below
```{r}
Confirmed_State_3_13_9_13_joined_long <- Confirmed_State_3_13_9_13_joined %>% 
              pivot_longer(-c(Province_State),
                            names_to = "Date", values_to = "Confirmed")
```

```{r, fig.width = 5, fig.height = 10}
# to adjust figure size the chuck heading should be: {r, fig.width = 5, fig.height = 10}

ggplot(Confirmed_State_3_13_9_13_joined_long, aes(x = Confirmed,  y = Province_State))  + 
    geom_point(aes(color = Date)) 
```
# Working with the time series data
Start by loading the most recent times series data for confirmed cases
```{r}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```
Check the table properties to make sure the data imported as we expected
```{r}
head(time_series_confirmed)
```
As noted above this data is in wide format. To convert to long format:
```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed")
```
Let’s look at the format of the data frame (tibble) now
```{r}
head(time_series_confirmed_long)
```
It would be convenient to have the confirmed cases and deaths in the same table. We can create another table with the deaths and then join the two tables. First let’s:
```{r}
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", 
               destfile = "data/time_series_covid19_deaths_global.csv")
# this is now saved in the /data folder 
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
# renames columns
```
As noted above this data is in wide format. To convert to long format
```{r}
time_series_deaths_long <- time_series_deaths %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)
```
# Joining the time series tables
To join 2 tables we need a common column in which each row is a unique name. This is often called the Key. None of the columns in this data set meet that criteria, but we can create a column to serve as our key by mergering the names for the Province_State, Country_Region and Date columns using unite
```{r}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
```
For the second table we can do the same and eliminate the columns that are redundant so that we just have the Key and Deaths columns
```{r}
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)
```
Now to join the tables. A full_join which means that the keys that are common to the tables will be joined and there will be counts for both Confirmed and Deaths. Where there are not matching values, returns NA for the one missing. select can be used to remove the key after the join since we don’t have further need for it.
```{r}
time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
              select(-Key)
head(time_series_long_joined)
```
Check to make sure neither the Cofirmed or Death counts have NA as a value (both tables have the same number of rows, but we should check to make sure)
```{r}
which(is.na(time_series_long_joined$Confirmed))

which(is.na(time_series_long_joined$Deaths))
```
There are no NA values. Now we are in good shape to proceed to using these table for graphing, but first reformat Date
```{r}
# load the lubridate library
library(lubridate)

time_series_long_joined$Date <- mdy(time_series_long_joined$Date)
```
Confirmed and Deaths are count data and there may be times when we want to plot both on the same graph. There are multiple ways to accomplish this and one is through creating a column using _pivot_long
```{r}
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")
head(time_series_long_joined_counts)

```
# Making Graphs from the time series data
To make a times series graph of the confirmed cases we need to summarize the Country date to count up the individual state data for the US.
```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Deaths")
```
Let’s look at the US data in the context of a few other countries
```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y")
```

Now several countries on the same graph
```{r}
time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% c("China","France","Italy", 
                                "Korea, South", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths")
```
We can use the alternative data frame with column Report_Type to show Confirmed and Deaths. Because the counts for Deaths are low relative to Confirmed a log scale works best here for seeing the pattern
```{r}
time_series_long_joined_counts %>% 
  group_by(Country_Region, Report_Type, Date) %>% 
  summarise(Counts = sum(Counts)) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = log2(Counts), fill = Report_Type, color = Report_Type)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Cases")
```

# Exercises

### 2. Revise the above example for joining tables using 6/13/2020 and 9/13/2020. Plot the data as a bar plot. Make sure your graph looks nice in the report by adjusting the height in the R code chunk header.

```{r}
# read in data for June, no need to rename the country column 
Confirmed_State_6_13 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/06-13-2020.csv")) %>%
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
# read in data for September 13 
Confirmed_State_9_13 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>%
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 

```
Find out if there are differences in the countries/ provinces in each data table
```{r}
setdiff(Confirmed_State_6_13$Province_State, Confirmed_State_9_13$Province_State)
# no differences (all states/provinces have reported cases at these points)
```
Merge the tables:
all of the states and territories in 6/13/2020 are found in 9/13/2020 a left_join or full_join can be used to merge the tables
```{r}
Confirmed_State_joined <- full_join(Confirmed_State_6_13,
      Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_joined)
tail(Confirmed_State_joined)
# this table now has the confirmed cases for 6/13 and 9/13 by state (US only)

```
Check for NA values
```{r}
which(is.na(Confirmed_State_joined))
```
We have columns Confirmed.x Country_Region.x, Confirmed.y Country_Region.y. So let’s rename the Confirmed columns and delete the Country_Region columns
```{r}
Confirmed_State_joined <- full_join(Confirmed_State_6_13,
      Confirmed_State_9_13, by = c("Province_State")) %>% 
      rename(Confirmed_6_13_2020 = "Confirmed.x", Confirmed_9_13_2020 = "Confirmed.y") %>% 
      select(-Country_Region.x, -Country_Region.y) 
head(Confirmed_State_joined)
```

```{r}
Confirmed_State_joined_long <- Confirmed_State_joined %>% 
              pivot_longer(-c(Province_State),
                            names_to = "Date", values_to = "Confirmed")
head(Confirmed_State_joined_long)
```

## Barplot

### 3. Add more informative x and y labels and a title to the above graph in Ex2.

```{r, fig.width = 5, fig.height = 3}

ggplot(data = Confirmed_State_joined_long, aes(x = Province_State, y = (Confirmed))) +
  geom_bar(stat = "identity", aes(color = Date)) +
  theme(axis.text.x = element_text(size = 6, angle = 90)) +
  labs(title = "US COIVD-19 Cases by State",
       x = "State/Province", 
       y = "Confirmed Cases")

```

### 4. Using the time series data. Make a plot of the total number of confirmed deaths per day worldwide

```{r}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```
convert to long format 
```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 
```
create another table with the deaths and then join the two tables
```{r}
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```
need to convert format 
```{r}
time_series_deaths_long <- time_series_deaths %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)
```
merge the two tables 
```{r}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
# same for other table 
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)

time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
              select(-Key)
head(time_series_long_joined)
```
make sure no counts are NA 
```{r}
which(is.na(time_series_long_joined$Confirmed))
```
load lubridate lib
```{r}
library(lubridate)
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)

```
long format 
```{r}
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")
head(time_series_long_joined_counts)
```
Plot deaths per day worldwide 
```{r, fig.width = 8, fig.height = 5}
#View(time_series_long_joined)
library(tidyverse)

p <- time_series_long_joined %>% 
  group_by(Date) %>% 
  summarise_at(c("Deaths"), sum) %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_jitter(alpha = 0.3) +
    ggtitle("Worldwide COVID-19 Deaths over time") +
  theme(axis.text.x = element_text(size =10, angle = 0)) 

p
```
save the graph as a picture file 
```{r}
png(filename = "WorldwideDeaths.png") 

p 

dev.off()

```

### 5. Use Mutate to make a new column with deaths/confirmed cases (Hint: this might be best done after joining the tables, but before pivoting longer).

```{r, fig.width = 8, fig.height = 5}
#library(tidyverse)
#?mutate

DC<- time_series_long_joined %>%
  filter(!is.na(Deaths), !is.na(Confirmed)) %>%
  mutate(DC = Deaths / Confirmed)

head(DC)
# DC is the new column (ration of Deaths: Confirmed Cases)
```

### 6. Plot US deaths/confirmed cases per day.

```{r, fig.width = 8, fig.height = 5}
DC %>%
  filter(Country_Region == "US") %>%
  ggplot(aes(x=Date, y=DC)) +
  geom_jitter() +
  #theme(axis.text.x = element_text(size = 2, angle = 90)) +
  labs(title = "US COVID-19 Death to Case ratio over time", 
       x = "Date",
       y = "Death to Case Ratio") +
  theme(axis.text.x = element_text(size =10, angle = 0)) 

```
### 7. Make a single graph with from the times series data with the 10 countries with the hightest death total.
```{r}
# use slice to get the top 10 death totals 

head(time_series_deaths_long)

top_death <- time_series_deaths_long %>%
  group_by(Country_Region) %>% #data by country
  summarise(sum_death = sum(Deaths)) %>%
  arrange(desc(sum_death)) %>% # most death to least
  slice(c(1:10)) # take only the top 10
  
head(top_death)
```

Single graph of countries with highest death toll 
```{r}
ggplot(data = top_death, mapping = aes(x = Country_Region, y = sum_death)) +
  geom_bar(stat = "identity") +
  labs(title = "Countries with Highest COVID-19 Deaths",
       x = "Country",
       y = "Total Death") 
```

### 8. Using Facet_Wraps make a graph from the times series data with the 10 countries with the hightest death total. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.

```{r, fig.width = 8, fig.height = 5}
top_death_timeseries <- time_series_long_joined %>%
  group_by(Country_Region) %>% # data by country
  filter(!is.na(Deaths)) %>%
  select(Deaths, Country_Region, Date) %>% # death by country over time
  summarise(Deaths, Date, Country_Region) %>%
  filter(Country_Region == top_death$Country_Region) #taking countries from the df with top total deaths 
# or manually 
# filter(Country_Region == c("Brazil", "France", "India", "Iran", "Italy", "Mexico", "Peru", "Spain", "United Kingdom", "US")) 

head(top_death_timeseries)
```

### 10. Apply a theme other than the one used in the Data Carpentries tutorial to the graph in Ex8.

```{r, fig.width = 8, fig.height = 5}
ggplot(data = top_death_timeseries, aes(x = Date,  y = Deaths, color = Country_Region)) + 
  geom_jitter() +
  facet_wrap(facets = vars(Country_Region)) +
  labs(title = "10 Countries with Highest COVID-19 Deaths Over Time") +
  theme_minimal()

# not sure why there are two trend lines in France and UK... 

```

### 9. Using Facet_Wraps make a graph from the times series data for confirmed cases for all US states and territories. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.

Hint: Instead of the global time series use the US time series data in the same folder -- time_series_covid19_deaths_US.csv and time_series_covid19_confirmed_US.csv 

Read in Data 
```{r}
# read in US confirmed time series data

US_time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")) 

head(US_time_series_confirmed)
#View(US_time_series_confirmed)
```

```{r}
# and the US deaths time series data

US_time_series_deaths <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")) 

head(US_time_series_deaths)

```
Convert to Long Format
```{r}
# Confirmed Data 
US_time_series_confirmed_long <- US_time_series_confirmed %>%
  pivot_longer(-c(UID, iso2,iso3, code3, FIPS, Admin2, Lat, Long_, Combined_Key, Province_State, Country_Region, Lat, Long_),
               names_to = c("Date"), values_to = c("Confirmed"))

head(US_time_series_confirmed_long)

# Death Data 
US_time_series_deaths_long  <- US_time_series_deaths %>%
  pivot_longer(-c(UID, iso2,iso3, code3, FIPS, Admin2, Lat, Long_, Combined_Key, Province_State, Country_Region, Lat, Long_, Population),
               names_to = "Date", values_to = "Deaths") 

head(US_time_series_deaths_long)
```
Subset Confirmed Case Data by State 
```{r}
US_time_series_confirmed_long <- US_time_series_confirmed_long %>%
  group_by(Province_State) %>%
  select(Province_State, Date, Confirmed)

head(US_time_series_confirmed_long)
tail(US_time_series_confirmed_long)
```
Find cumulative sum of cases by State 
```{r}
US_Sum_State <-US_time_series_confirmed_long %>%
  group_by(Province_State, Date) %>%
  summarise(Cases = sum(Confirmed))

head(US_Sum_State)
```
Plotting cumulative sum of US covid-19 Cases by state over time 
```{r, fig.width = 10, fig.height = 5}
ggplot(data = US_Sum_State, aes(x = Date, y = Cases, color = Province_State)) +
  geom_area() +
  facet_wrap(facets = vars(Province_State)) +
  ggtitle("US COVID-19 Cases by State") + 
  theme(strip.text.x = element_text(size = 8)) 

```

Application written in R [@RCoreTeam] using the Shiny framework [@Chang2015] and the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University [@Dong2020]. 

# References 


