---
title: "lab 6 exercises"
author: "Margaret Hilliard"
date: "9/28/2020"
output: html_document
bibliography: /Users/margarethilliard/Desktop/Evolutionary_Genomics/genomics-course/data/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```
load libraries
```{r}
library(tidyverse)
```

# Exercises

### 1. For the above graph “World COVID-19 Confirmed case” summarize the counts for each Country on the graph and update the graph to 9/26/2020. You may need to adjust the size of the points.

import case data from 9/26/20
```{r}
World_report_09_26_2020 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_") %>%
  group_by(Country_Region) %>%
  summarize(Confirmed = sum(Confirmed), Lat, Long)

# Lat = mean(Lat), Long = mean(Long)
head(World_report_09_26_2020)
```
```{r}
world_breaks <- c(1, 1000, 10000, 100000, 1000000, )
ggplot(World_report_09_26_2020, aes(x = Long, y = Lat, size = Confirmed/100,000)) +
    borders("world", colour = NA, fill = "grey90") +
    theme_bw() +
    geom_point(shape = 21, color='red', fill='red', alpha = 0.3) +
    scale_size_continuous(name = "Cases (x100,000)", range = c(1,4)) +
    labs(title = 'World COVID-19 Confirmed cases',x = '', y = '') +
    theme(legend.position = "right")

```

### 2. Update Anisa Dhana’s graph layout of the US to 9/26/2020. You may need to adjust the size of the points.

```{r}

US_report_09_26_2020 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_") %>%
  filter(Country_Region == "US") %>%
  filter (!Province_State %in% c("Alaska","Hawaii", "American Samoa",
                  "Puerto Rico","Northern Mariana Islands", 
                  "Virgin Islands", "Recovered", "Guam", "Grand Princess",
                  "District of Columbia", "Diamond Princess")) %>% 
   filter(Lat > 0)

head(US_report_09_26_2020)

ggplot(US_report_09_26_2020, aes(x = Long, y = Lat, size = Confirmed/1000)) +
    borders("state", colour = "black", fill = "grey90") +
    theme_bw() +
    geom_point(shape = 21, color='purple', fill='purple', alpha = 0.5) +
    labs(title = 'COVID-19 Confirmed Cases in the US', x = '', y = '',
        size="Cases (x1000))") +
    theme(legend.position = "right") +
    coord_fixed(ratio=1.5)
```

```{r}
mybreaks <- c(1, 100, 1000, 10000, 100000)

ggplot(US_report_09_26_2020, aes(x = Long, y = Lat, size = Confirmed)) +
    borders("state", colour = "white", fill = "grey90") +
    geom_point(aes(x = Long, y = Lat, size = Confirmed, color = Confirmed), stroke=F, alpha=0.5) +
    scale_size_continuous(name = "Cases", range = c(1,15), 
                        breaks = mybreaks, labels = c("1-99",
                        "100-999", "1,000-9,999", "10,000-99,999", "100,000+")) +
    scale_color_viridis_c(option = "viridis", name = "Cases", 
                          breaks = mybreaks, labels = c("1-99",
                        "100-999", "1,000-9,999", "10,000-99,999", "100,000+"))  +
# Cleaning up the graph
  
  theme_void() + 
    guides( colour = guide_legend()) +
    labs(title = "Anisa Dhana's lagout for COVID-19 Confirmed Cases in the US") +
    theme(
      legend.position = "bottom",
      text = element_text(color = "#22211d"),
      plot.background = element_rect(fill = "#ffffff", color = NA), 
      panel.background = element_rect(fill = "#ffffff", color = NA), 
      legend.background = element_rect(fill = "#ffffff", color = NA)
    ) +
    coord_fixed(ratio=1.5)


```

### 3. Update the above graph “Number of Confirmed Cases by US County” to 9/26/2020 and use a different color scheme or theme

```{r}
library(RColorBrewer)
# To display only colorblind-friendly brewer palettes, specify the option colorblindFriendly = TRUE as follow:
# display.brewer.all(colorblindFriendly = TRUE)
# Get and format the covid report data
report_09_26_2020 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  unite(Key, Admin2, Province_State, sep = ".") %>% 
  group_by(Key) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Key = tolower(Key))
```

get and format the map data
```{r}
us <- map_data("state")
counties <- map_data("county") %>% 
  unite(Key, subregion, region, sep = ".", remove = FALSE)
# Join the 2 tibbles
state_join <- left_join(counties, report_09_26_2020, by = c("Key"))

# sum(is.na(state_join$Confirmed))
```


```{r}
ggplot(data = us, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  # Add data layer
  borders("state", colour = "black") +
  geom_polygon(data = state_join, aes(fill = Confirmed)) +
  scale_fill_gradientn(colors = brewer.pal(n = 10, name = "OrRd"),
                       breaks = c(1, 10, 100, 1000, 10000, 100000),
                       trans = "log10", na.value = "White") +
  ggtitle("Number of Confirmed Cases by US County") +
  theme_bw() 
```

### 4. Make an interactive plot using a state of your chosing using a theme different from used in the above exammples.

```{r}
TX_daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  filter(Province_State == "Texas") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2))

us <- map_data("state")

tx_us <- subset(us, region == "texas")

counties <- map_data("county")

tx_county <- subset(counties, region == "texas")

state_join <- left_join(tx_county, TX_daily_report, by = c("subregion" = "Admin2"))
```

borders("state", colour = "white", fill = "grey90") +
```{r}
# plot state map
ggplot(data = tx_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "white") +
    scale_fill_gradientn(colors = brewer.pal(n = 8, name = "Spectral"),
                         trans = "log10") +
  labs(title = "COVID-19 Confirmed Cases in Texas",
       x = "", 
       y = "")
```

```{r}
library(plotly)
ggplotly(
  ggplot(data = tx_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
    scale_fill_gradientn(colours = 
                         wes_palette("Chevalier1", type = "continuous")) +
  ggtitle("COVID-19 Cases in TX") +
# Cleaning up the graph
  labs(x=NULL, y=NULL) +
  theme(panel.border = element_blank()) +
  theme(panel.background = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(axis.text = element_blank())
)

?scale_fill_gradientn
```

Application written in R [@RCoreTeam] using the Shiny framework [@Chang2015] and the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University [@Dong2020]. 

# References 
